The core transcriptome of mammalian placentas and the divergence of expression with placental shape

### 2017-03-08 manuscript/manuscript.Rnw
use ortholog pairs and describe OMA analysis

Single-copy (1-to-1) orthologs in were determined using Ensembl v.80
gene trees via Biomart using the biomaRt R package
\cite{Durinck.ea2009:Mappingidentifiersforintegration}. %
A total of \Sexpr{combined.fpkm[species!="homo sapiens" & mean_fpkm >=
  1,length(unique(human_name))]} single-copy orthologous genes were
expressed in at least one species. We also obtained orthology from the
perspective of each species to all other species using biomaRt package
and Ensembl v.80 database to determine the percentage of highly
expressed genes which had an orthology of one-to-one, one-to-many,
many-to-many, or novel/unknown.

Significant shifts in expression levels between groups of species were
calculated using Student's t test on single-copy orthologs using the mean of
each species to avoid large biases towards samples with $n \gg 1$ and
considered significant if the \ac{FDR} as estimated by the Benjamini
and Hochberg procedure \cite{Benjamini1995:bh_procedure} was
$\le 0.05$. For \human and \mouse lineage specific calculations, a
type-III anova was calculated using
lme4~\cite{Bates.ea2015:FittingLinearMixed-EffectsModels} with each
species having separate random effects. Shifts in expression levels
are reported with $\log_2$ Fold-Change (FC), rounded to the nearest
integer.

@@

Supplemental \cref{fig:transcriptome_statistics} shows the RNA-Seq read and
alignment statistics for all species. For transcriptomes sequenced
with 76 bp %
%
(\Sexpr{species(star.logs.wide[input.read.length > 70 & input.read.length < 80,unique(species)])}),%
%
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,max(as.numeric(input.reads))]}
with an average of \Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,mean(`% good reads`)]}\% of reads 
mapping to the respective reference genome (\cref{tab:ref_genome_aligned_species}). 
%
For transcriptomes sequenced with 150 bp %
% The mismatch here is because we're reporting trimmed reads here, not
% full length reads
(\Sexpr{species(star.logs.wide[input.read.length > 120 & input.read.length < 160,unique(species)])}),
%
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length > 120 & input.read.length < 160,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length  > 120 & input.read.length < 160,max(as.numeric(input.reads))]}
%
with an average of \Sexpr{star.logs.wide[input.read.length > 120 & input.read.length < 160,mean(`% good reads`)]}\% %
of reads mapping to the
respective reference genome (Supplemental \cref{tab:ref_genome_aligned_species}).
%
%
For transcriptomes sequenced with 100 bp %
% The mismatch here is because we're reporting trimmed reads here, not
% full length reads
(\Sexpr{species(star.logs.wide[input.read.length > 80 & input.read.length < 120,unique(species)])}),
%
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length > 80 & input.read.length < 120,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length  > 80 & input.read.length < 120,max(as.numeric(input.reads))]}
%
with an average of \Sexpr{star.logs.wide[input.read.length > 80 & input.read.length < 120,mean(`% good reads`)]}\% %
of reads mapping to the
respective reference genome (Supplemental \cref{tab:ref_genome_aligned_species}).
%
%
For transcriptomes sequenced with 35 bp %
% The mismatch here is because we're reporting trimmed reads here, not
% full length reads
(\Sexpr{species(star.logs.wide[input.read.length < 40 ,unique(species)])}),
%
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length < 40 ,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length  < 40  ,max(as.numeric(input.reads))]}
%
with an average of %
\Sexpr{star.logs.wide[input.read.length < 40, mean(`% good reads`)]}\%
%
of reads mapping to the respective reference genome (Supplemental
\cref{tab:ref_genome_aligned_species}). Read length was not a
significant predictor of unmapped read percentage (Supplemental
\cref{fig:transcriptome_reads_vs_read_length}). %
%
Between 20\% and 40\% of all highly expressed genes in each species
were single-copy orthologs with all other analyzed species, with the
remainder being mostly one-to-many or many-to-many in comparison to at
least one other species (Supplemental
\cref{tab:orthologs_in_each_species}).
%
The expression of
all genes with \ac{FPKM} $\ge 50$ is shown in Supplemental
\cref{tab:all_ateles_fusciceps_placenta,tab:all_homo_sapiens_placenta,tab:all_pan_paniscus_placenta,tab:all_mus_musculus_placenta,tab:all_spalax_galili_placenta,tab:all_spalax_carmeli_placenta,tab:all_bos_taurus_placenta,tab:all_ovis_aries_placenta,tab:all_sus_scrofa_placenta,tab:all_equus_caballus_placenta,tab:all_canis_familiaris_placenta,tab:all_loxodonta_africana_placenta,tab:all_dasypus_novemcinctus_placenta,tab:all_monodelphis_domestica_placenta}.

@@

print(ggplot(cptl[FPKM > 0.1,],aes(x=human_name,y=FPKM))+
      geom_boxplot(outlier.shape=NA)+
      scale_y_log10(labels=trans_format("log10", math_format(10^.x)))+
      geom_point(aes(color=Species,shape=Species))+
      scale_shape_manual("Species",
                         guide=guide_legend(#nrow=1,
                                            keywidth=unit(3,"mm"),
                                            keyheight=unit(3,"mm"),
                                            label.theme=element_text(face="italic",
                                                                     angle=0,
                                                                     size=7),
                                            title.position="top"
                                            ),
                         values=0:18)+
      # scale_color_discrete(NULL)+
      coord_flip()+
      xlab("Human Ortholog Name")+
      ylab("Gene Expression (FPKM)")+
      theme_bw() +
      theme(text=element_text(size=9),
            legend.position=c(0.99,0.41),
            axis.text.y=element_text(face="italic",size=7),
            legend.margin=unit(0,"mm"),
            legend.justification=c(1,1),
            plot.margin=unit(c(0,0,0,0),"mm")
            ))
@ 
\caption{Core Placenta Transcriptome.
  \Sexpr{core.placenta.transcriptome.long[housekeeping==FALSE,length(unique(human_name))]}
  single-copy \human ortholog genes expressed (\ac{FPKM}$ \ge 10$) in the
  placenta of all 13 eutherians studied with housekeeping genes
  \cite{Eisenberg.ea2013:Humanhousekeepinggenesrevisited} removed
  ordered (from top to bottom) by median expression. If we also require
  expression (\ac{FPKM}$ \ge 10$) in the marsupial, \opossum, there are 
  \Sexpr{cpt.distribution.shape[`Expression (FPKM)` > 10 &
    `Min. Exp. in` == '14 of 14 Species',max(num_genes)]} genes
  which meet this criteria.
  \Cref{fig:core_placenta_transcriptome_shape} shows the shape of the
  distribution of the number of genes in the core placenta
  transcriptome if different \ac{FPKM} and species restriction criteria are
  utilized.}
\label{fig:core_placenta_transcriptome}
\end{figure}

@@

              legend.position=c(0.99,0.99),
              legend.margin=unit(0,"mm"),
              legend.justification=c(1,1),

@@

  shows the number of single-copy orthologous genes included in the core

@@

  genes would be included if the expression of a set of single-copy orthologs of all 13 Eutherians must be at
  least 10 FPKM, %

@@

  single-copy orthologs would be included if 13 of 14 species have an

@@

The
\Sexpr{core.placenta.transcriptome.long[housekeeping==FALSE,length(unique(human_name))]}
genes comprising the core placenta transcriptome are shown in
\cref{fig:core_placenta_transcriptome} ordered by their median
transcription level in \ac{FPKM}. We considered a gene to be core to
the placenta transcriptome if a single-copy ortholog of that gene was
expressed with $\text{FPKM} \ge 10$ in all eutherian species studied
(excluding the metatherian, \opossum) and was not a human housekeeping
gene according to
\citet{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}. Genes which
meet this criteria but are housekeeping genes are depicted in
Supplemental \cref{fig:core_placenta_transcriptome_housekeeping}.
\Cref{fig:core_placenta_transcriptome_shape} shows the number of genes
which would be included in the core placenta set if the \ac{FPKM}
cutoff was set differently; for example, choosing an \ac{FPKM} cutoff
of 1 yields %
\Sexpr{cpt.distribution.shape[`Expression (FPKM)`>=1 & `Min. Exp. in`
  == '14 of 14 Species',max(num_genes)]} %
genes in the core transcriptome. %
In contrast to the shape of the number of genes included in the core
placenta transcriptome, the set of \human housekeeping genes expressed
all placentas is flatter (Supplemental
\cref{fig:core_housekeeping_transcriptome_shape}), and includes more
of the total fraction of genes (%
%
\Sexpr{core.placenta.transcriptome.long[housekeeping==TRUE,length(unique(human_name))]/
  core.placenta.transcriptome.long[,length(unique(human_name))]} %
%
%
vs %
\Sexpr{core.placenta.transcriptome.long[housekeeping==FALSE,length(unique(human_name))]/
  core.placenta.transcriptome.long[,length(unique(human_name))]}%
%
). An alternative approach to defining
housekeeping genes is to use the tissue specificity
index~\cite{Yanai.ea2005:Genome-widemidrangetranscriptionprofiles}.
Supplemental \cref{tab:tissue_specificity_index_of_plaenta_core_genes}
lists the tissue specificity index for all housekeeping and
non-housekeeping genes; housekeeping genes expressed in all placentas
have a lesser median tissue specificity index than non-housekeeping
genes (Supplemental
\cref{fig:tissue_specificity_index_with_housekeeping}).

Supplemental \cref{tab:core_placenta_go_terms} lists the significant
enrichment of \ac{GO} terms in the core placenta transcriptome;
notable terms include those involved in cytoskeletal structure,
protein transport, membrane protein structure, and inflammation
resolution. Increasing the permissiveness of the core placenta
transcriptome by only requiring 10 of 14 species to express single-copy
orthologs with FPKM $>10$ and results in the same terms also being
significantly enriched, along with additional terms (Supplemental
\cref{tab:core_placenta_go_terms}). Supplemental
\cref{tab:core_placenta_pathway_analysis} shows the pathways with
significant and suggestive enrichment in the core placenta
transcriptome.

If we use gene trees to define the core placenta transcriptome instead
of single-copy orthologs (where the mean expression for genes in the
tree which are present must be $\text{FPKM} \ge 10$), we find
\Sexpr{cpttl[,length(unique(tree_genes))]} unique gene trees which
include nearly all of the genes present in the core placenta
transcriptome (Supplemental
\cref{fig:core_placenta_tree_transcriptome,fig:core_placenta_tree_transcriptome_housekeeping}).

>>>>

Orthologous pairs between \human and all other species were identified
using the \ac{OMA} database
\cite{Altenhoff.ea2015:OMAorthologydatabasein}. In cases where a gene
did not have a human ortholog pair identified by \ac{OMA}, Ensembl v.80
one2one orthologs obtained from Biomart using the biomaRt R package
\cite{Durinck.ea2009:Mappingidentifiersforintegration} %
were used to select the appropriate orthologous pair. A total of
\Sexpr{combined.fpkm[,list(n=length(gene_id),
  max.fpkm=max(mean_fpkm)),by=human_name][n==14 & max.fpkm >
  10,length(human_name)]} ortholog pair sets in all 14 species were
expressed (\ac{FPKM} $\ge 10$) in at least one species. We also
obtained orthology from the perspective of each species to all other
species using biomaRt package and Ensembl v.80 database to determine
the percentage of highly expressed genes which had an orthology of
one-to-one, one-to-many, many-to-many, or novel/unknown. For future
analyses by other investigators, we have also provided the human gene
which most closely aligns to each expressed annotated
protein-generating sequence of all species analyzed using DIAMOND
\cite{Buchfink.ea2015:Fastandsensitiveprotein}.

Significant shifts in expression levels between groups of species were
calculated using Student's t test on ortholog pairs using the mean of
each species to avoid large biases towards samples with $n \gg 1$ and
considered significant if the \ac{FDR} as estimated by the Benjamini
and Hochberg procedure \cite{Benjamini1995:bh_procedure} was
$\le 0.05$. For \human and \mouse lineage specific calculations, a
type-III anova was calculated using
lme4~\cite{Bates.ea2015:FittingLinearMixed-EffectsModels} with each
species having separate random effects. Shifts in expression levels
are reported with $\log_2$ Fold-Change (FC), rounded to the nearest
integer.

@@

Supplemental \cref{fig:transcriptome_statistics} shows the RNA-Seq read and
alignment statistics for all species. For transcriptomes sequenced
with 76 bp %
%
(\Sexpr{species(star.logs.wide[input.read.length > 70 & input.read.length < 80,unique(species)])}),%
%
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,max(as.numeric(input.reads))]}
with an average of \Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,mean(`% good reads`)]}\% of reads 
mapping to the respective reference genome (\cref{tab:ref_genome_aligned_species}). 
%
For transcriptomes sequenced with 150 bp %
% The mismatch here is because we're reporting trimmed reads here, not
% full length reads
(\Sexpr{species(star.logs.wide[input.read.length > 120 & input.read.length < 160,unique(species)])}),
%
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length > 120 & input.read.length < 160,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length  > 120 & input.read.length < 160,max(as.numeric(input.reads))]}
%
with an average of \Sexpr{star.logs.wide[input.read.length > 120 & input.read.length < 160,mean(`% good reads`)]}\% %
of reads mapping to the
respective reference genome (Supplemental \cref{tab:ref_genome_aligned_species}).
%
%
For transcriptomes sequenced with 100 bp %
% The mismatch here is because we're reporting trimmed reads here, not
% full length reads
(\Sexpr{species(star.logs.wide[input.read.length > 80 & input.read.length < 120,unique(species)])}),
%
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length > 80 & input.read.length < 120,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length  > 80 & input.read.length < 120,max(as.numeric(input.reads))]}
%
with an average of \Sexpr{star.logs.wide[input.read.length > 80 & input.read.length < 120,mean(`% good reads`)]}\% %
of reads mapping to the
respective reference genome (Supplemental \cref{tab:ref_genome_aligned_species}).
%
%
For transcriptomes sequenced with 35 bp %
% The mismatch here is because we're reporting trimmed reads here, not
% full length reads
(\Sexpr{species(star.logs.wide[input.read.length < 40 ,unique(species)])}),
%
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length < 40 ,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length  < 40  ,max(as.numeric(input.reads))]}
%
with an average of %
\Sexpr{star.logs.wide[input.read.length < 40, mean(`% good reads`)]}\%
%
of reads mapping to the respective reference genome (Supplemental
\cref{tab:ref_genome_aligned_species}). Read length was not a
significant predictor of unmapped read percentage (Supplemental
\cref{fig:transcriptome_reads_vs_read_length}). %
%
Between 20\% and 40\% of all highly expressed genes in each species
were ortholog pairs with all other analyzed species, with the
remainder being mostly one-to-many or many-to-many in comparison to at
least one other species (Supplemental
\cref{tab:orthologs_in_each_species}).
%
The expression of
all genes with \ac{FPKM} $\ge 50$ is shown in Supplemental
\cref{tab:all_ateles_fusciceps_placenta,tab:all_homo_sapiens_placenta,tab:all_pan_paniscus_placenta,tab:all_mus_musculus_placenta,tab:all_spalax_galili_placenta,tab:all_spalax_carmeli_placenta,tab:all_bos_taurus_placenta,tab:all_ovis_aries_placenta,tab:all_sus_scrofa_placenta,tab:all_equus_caballus_placenta,tab:all_canis_familiaris_placenta,tab:all_loxodonta_africana_placenta,tab:all_dasypus_novemcinctus_placenta,tab:all_monodelphis_domestica_placenta}.

@@

print(ggplot(cptl[FPKM > 0.1,],aes(x=human_name,y=FPKM))+
      geom_boxplot(outlier.shape=NA)+
      scale_y_log10(labels=trans_format("log10", math_format(10^.x)))+
      geom_point(aes(color=Species,shape=Species))+
      scale_shape_manual("Species",
                         guide=guide_legend(#nrow=1,
                                            keywidth=unit(3,"mm"),
                                            keyheight=unit(3,"mm"),
                                            label.theme=element_text(face="italic",
                                                                     angle=0,
                                                                     size=7),
                                            title.position="top"
                                            ),
                         values=0:18)+
      # scale_color_discrete(NULL)+
      coord_flip()+
      xlab("Human Ortholog Name")+
      ylab("Gene Expression (FPKM)")+
      theme_bw() +
      theme(text=element_text(size=9),
            legend.position=c(0.99,0.41),
            axis.text.y=element_text(face="italic",size=7),
            legend.margin=unit(0,"mm"),
            legend.justification=c(1,1),
            plot.margin=unit(c(0,0,0,0),"mm")
            ))
@ 
\caption{Core Placenta Transcriptome.
  \Sexpr{core.placenta.transcriptome.long[housekeeping==FALSE,length(unique(human_name))]}
  ortholog pairs of \human genes expressed (\ac{FPKM}$ \ge 10$) in the
  placenta of all 13 eutherians studied with housekeeping genes
  \cite{Eisenberg.ea2013:Humanhousekeepinggenesrevisited} removed
  ordered (from top to bottom) by median expression. If we also require
  expression (\ac{FPKM}$ \ge 10$) in the marsupial, \opossum, there are 
  \Sexpr{cpt.distribution.shape[`Expression (FPKM)` > 10 &
    `Min. Exp. in` == '14 of 14 Species',max(num_genes)]} genes
  which meet this criteria.
  \Cref{fig:core_placenta_transcriptome_shape} shows the shape of the
  distribution of the number of genes in the core placenta
  transcriptome if different \ac{FPKM} and species restriction criteria are
  utilized.}
\label{fig:core_placenta_transcriptome}
\end{figure}

@@

              legend.position=c(0.01,0.01),
              legend.margin=margin(0,0,0,0,unit="mm"),
              legend.spacing=unit(0,"mm"),
              legend.justification=c(0,0),

@@

  shows the number of ortholog pairs included in the core

@@

  genes would be included if the expression of a set of ortholog pairs of all 13 Eutherians must be at
9  least 10 FPKM, %

@@

  ortholog pairs would be included if 13 of 14 species have an

@@

The
\Sexpr{core.placenta.transcriptome.long[housekeeping==FALSE,length(unique(human_name))]}
genes comprising the core placenta transcriptome are shown in
\cref{fig:core_placenta_transcriptome} ordered by their median
transcription level in \ac{FPKM}. We considered a gene to be core to
the placenta transcriptome if an ortholog pair of that gene was
expressed with $\text{FPKM} \ge 10$ in all eutherian species studied
(excluding the metatherian, \opossum) and was not a human housekeeping
gene according to
\citet{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}. Genes which
meet this criteria but are housekeeping genes are depicted in
Supplemental \cref{fig:core_placenta_transcriptome_housekeeping}.
\Cref{fig:core_placenta_transcriptome_shape} shows the number of genes
which would be included in the core placenta set if the \ac{FPKM}
cutoff was set differently; for example, choosing an \ac{FPKM} cutoff
of 1 yields %
\Sexpr{cpt.distribution.shape[`Expression (FPKM)`>=1 & `Min. Exp. in`
  == '14 of 14 Species',max(num_genes)]} %
genes in the core transcriptome. %
In contrast to the shape of the number of genes included in the core
placenta transcriptome, the set of \human housekeeping genes expressed
all placentas is flatter (Supplemental
\cref{fig:core_housekeeping_transcriptome_shape}), and includes more
of the total fraction of genes (%
%
\Sexpr{core.placenta.transcriptome.long[housekeeping==TRUE,length(unique(human_name))]/
  core.placenta.transcriptome.long[,length(unique(human_name))]} %
%
%
vs %
\Sexpr{core.placenta.transcriptome.long[housekeeping==FALSE,length(unique(human_name))]/
  core.placenta.transcriptome.long[,length(unique(human_name))]}%
%
). An alternative approach to defining
housekeeping genes is to use the tissue specificity
index~\cite{Yanai.ea2005:Genome-widemidrangetranscriptionprofiles}.
Supplemental \cref{tab:tissue_specificity_index_of_plaenta_core_genes}
lists the tissue specificity index for all housekeeping and
non-housekeeping genes; housekeeping genes expressed in all placentas
have a lesser median tissue specificity index than non-housekeeping
genes (Supplemental
\cref{fig:tissue_specificity_index_with_housekeeping}).

Supplemental \cref{tab:core_placenta_go_terms} lists the significant
enrichment of \ac{GO} terms in the core placenta transcriptome;
notable terms include those involved in cytoskeletal structure,
protein transport, membrane protein structure, and inflammation
resolution. Increasing the permissiveness of the core placenta
transcriptome by only requiring 10 of 14 species to express an
ortholog pair with FPKM $>10$ and results in the same terms also being
significantly enriched, along with additional terms (Supplemental
\cref{tab:core_placenta_go_terms}). Supplemental
\cref{tab:core_placenta_pathway_analysis} shows the pathways with
significant and suggestive enrichment in the core placenta
transcriptome.

If we use gene trees to define the core placenta transcriptome instead
of ortholog pairs (where the mean expression for genes in the tree
which are present must be $\text{FPKM} \ge 10$), we find
\Sexpr{cpttl[,length(unique(tree_genes))]} unique gene trees which
include nearly all of the genes present in the core placenta
transcriptome (Supplemental
\cref{fig:core_placenta_tree_transcriptome,fig:core_placenta_tree_transcriptome_housekeeping}).
